---
title: "Exploring VI in Toy Realms"
author: "Ed Phillips"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring VI in Toy Realms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
library(vi.foundry)
```

## Scientific motivation

The empirical tests in the vi-foundry are blocked on data (Review Items 4–6:
gene-category data, island-bird morphology, hominin/economic time series).
The formal model is a theoretical ODE that cannot fail — it simulates VI's
equation and produces VI's prediction by construction. Between these lies a
gap: a layer that lets a reader *explore* the consequences of VI across
parameter space and hypothetical substrates, without claiming empirical test.

The **toy realms** fill this gap. Each realm is a self-contained "what if"
world with a simulator, a set of dials, and a visualization. They ask: *if VI
were true, what would we expect to see in worlds we have not measured?* The
answer sharpens the predictions and, when the data arrives, makes the
empirical tests more discriminating.

This is **not** empirical testing. The realms do not source new data, do not
claim to corroborate VI, and do not replace the blocked empirical work. Each
realm names the experiment that would convert it from speculative to
empirical. See `docs/review/toy-realms-plan.md` for the full execution plan.

## Method overview

| Realm | What it explores | Reuses | Status |
|-------|-----------------|--------|--------|
| 1. Genome reduction | The threshold gate as a control parameter | `threshold_model()` | Built |
| 2. Irreversibility | Hysteresis loop area vs. cusp parameter `a` | `cusp_hysteresis_check()` | Planned |
| 3. *Homo* inversion | Diversity-dependence sign flip | `generate_autocatalytic_set()` | Planned |
| 4. Cross-kingdom | When model transfer beats sign-only transfer | `empirical_formal_model()` | Planned |

## Realm 1 — The threshold gate

### What it explores

The threshold gate is the heart of VI's biphasic prediction. The formal model
([`threshold_model()`](../reference/threshold_model.html)) integrates:

$$\frac{dC_i}{dt} = -\lambda \cdot M(t) \cdot C_i \cdot \mathbb{1}(d_i < \theta)$$

where $C_i$ is the retention probability of trait $i$, $M(t) = M_0 e^{-\alpha t}$
is the decaying niche-demand mismatch, $d_i$ is the trait's integration depth,
and $\theta$ is the protection threshold. Traits with $d_i \geq \theta$ are
protected (retention = 1.0); traits with $d_i < \theta$ are shed
(retention $\to 0$).

The **biphasic signal** is the gate itself: protected traits retain at 1.0
while unprotected traits shed to ~0, producing a sharp contrast
(`threshold_biphasicity` ≈ 1.0). This is the real biphasic signature — not
the temporal displacement ratio (math-review Issue 3, resolved).

This realm makes the gate *visible* as a control parameter. By sweeping
$\theta$ across a grid, we can watch the gate open (biphasicity → 1) when
$\theta$ separates protected from unprotected traits, and close (biphasicity
→ 0) when $\theta$ is below the minimum depth (all protected) or above the
maximum depth (all unprotected).

### Worked example

Consider a hypothetical organism with five traits at integration depths
`r paste(c(0, 1, 2, 3, 5), collapse = ", ")`. We sweep $\theta$ from 0 to 6:

```{r sweep}
result <- sweep_threshold(
  depths = c(0, 1, 2, 3, 5),
  theta_grid = seq(0, 6, by = 0.25)
)
```

The sweep returns a data frame with one row per $\theta$ value:

```{r sweep-head}
head(result$values$sweep, 8)
```

The peak biphasicity is `r result$values$peak_biphasicity` at
$\theta = `r result$values$peak_theta`$. This is where the gate is maximally
open — the cleanest separation between protected and unprotected traits.

```{r plot-gate, fig.alt = "Threshold gate: biphasicity vs theta. The gate opens at theta > 0, stays at 1.0 through the interior, and closes at theta > 5."}
plot_threshold_gate(result)
```

The shaded region is the "gate open" zone — the parameter range where VI's
biphasic prediction is active. The vertical dashed lines mark each trait's
depth; the gate composition changes at each one.

### What this teaches

1. **The biphasic signal IS the gate.** The `threshold_biphasicity` metric
   captures the protected-vs-unprotected contrast. It is 1.0 whenever both
   classes are non-empty, and 0 when one class is empty. This is the signal
   the math-review (Issue 3) identified as the real biphasic signature — not
   the temporal displacement ratio.

2. **The gate is a step function.** The biphasicity rises sharply from 0 to 1
   as $\theta$ crosses the minimum depth, stays at 1 through the interior,
   and falls sharply to 0 as $\theta$ crosses the maximum depth. The gate is
   binary — it is either open or closed.

3. **The R6 method-misspecification is visceral.** A cross-sectional
   regression across organisms with *different* $\theta$ values would see
   noise, not a logistic — which is exactly why T3 fails on real
   endosymbiont data. Different lineages have different thresholds (different
   dependency architectures, different parasitism levels), so pooling them
   cross-sectionally destroys the within-lineage biphasic signal.

### Skewing the dependency architecture

The gate's shape depends on the dependency architecture. A uniform
architecture (`c(0, 1, 2, 3, 5)`) opens the gate gradually. A skewed
architecture — many shallow traits, few deep — changes the composition:

```{r skewed}
skewed <- sweep_threshold(
  depths = c(0, 0, 0, 1, 5, 5),
  theta_grid = seq(0, 6, by = 0.25)
)
plot_threshold_gate(skewed)
```

The gate still opens and closes, but the protected/unprotected counts change
differently. At $\theta = 0.5$, three traits are unprotected (the three at
depth 0) and three are protected — a different composition from the uniform
case.

### The experiment that would fill this realm

A real organism with a known dependency architecture (gene categories +
integration-depth scores) and a measured retention trajectory across a
parasitism/commitment gradient. The $\theta$-sweep would then be calibrated
against real data, and the gate's $\theta$ would be an empirical estimate,
not a dial. The foundry's `empirical_formal_model()` (Remark R7) is a step
in this direction — it fits the retention matrix to estimate $dep$ and $para$
effects — but the full gate calibration requires gene-category data (Item 4).

## Limitations

The toy realms are **theoretical exploration, not empirical test**:

- They do not source new data.
- They do not claim to corroborate VI.
- They do not replace the blocked empirical work (Items 4–6).
- The sweep uses the default ODE parameters ($\lambda = 0.15$, $M_0 = 10$,
  $\alpha = 0.05$, $t = 100$); with different parameters, the unprotected
  traits may not fully shed, and the biphasicity would be less than 1.0 in
  the interior. The gate shape (0 → contrast → 0) is robust, but the
  plateau height is parameter-dependent.

## Interpretation guide

| What you see | What it means |
|-------------|---------------|
| Biphasicity = 0 at $\theta = 0$ | All traits protected — no contrast (gate closed) |
| Biphasicity = 1 in the interior | Gate open — clean separation between protected and unprotected |
| Biphasicity = 0 at $\theta > \max(d)$ | All traits unprotected — no contrast (gate closed) |
| Vertical dashed lines | Trait depths — gate composition changes here |
| Peak $\theta$ | The threshold that maximally separates the two classes |

The key insight: **the threshold gate is the biphasic signal**. When the
gate is open (both classes exist), VI's biphasic prediction is active. When
the gate is closed (one class empty), there is no biphasic contrast. This is
why T3's cross-sectional regression fails — it pools organisms with different
gates, averaging the contrast away.
